A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-19 tex/codatadp.tex
more

%\renewcommand{\vec}[1]{\mathbf{#1}}
\renewcommand{\vec}[1]{\bar #1}
\newcommand{\Ec}{\mathsf{E}}
\newcommand{\Fc}{\mathsf{F}}
\newcommand{\Gc}{\mathsf{G}}
\newcommand{\Ac}{\mathsf{A}}
\newcommand{\Dc}{\mathsf{D}}
\newcommand{\Rc}{\mathsf{R}}
\newcommand{\vrange}{\mathsf{range}}
\newcommand{\vdom}{\mathsf{dom}}
\newcommand{\tEc}{\mathcal{T}(\Ec)}
\newcommand{\tcEc}{\mathcal{T}^\ast(\Ec)}
\newcommand{\rn}[1]{\textsf{\small #1}}
\newcommand{\cvc}{\textsc{cvc}{\small 4}\xspace}
\newcommand{\teq}{\approx}
\newcommand{\tneq}{\not\teq}
\newcommand{\rem}[1]{\textcolor{red}{[#1]}}
%\newcommand{\is}[1]{is\text{-} #1}
\newcommand{\is}[1]{\const{is#1}}
\newcommand{\ror}{\quad \parallel \quad}
\newcommand{\tpath}[2]{\mathcal{P}_{ #2 \rightarrow \_ }( #1 )}
\newcommand{\ttpath}[3]{\mathcal{P}_{ #2 \rightarrow #3 }( #1 )}
\newcommand{\interp}[1]{[\![ #1 ]\!]}
\newcommand{\dpath}[3]{\delta^{#1}_{#2}( #3 )}
\newcommand{\ec}[1]{ [ #1 ] }
\newcommand{\M}{\mathcal{M}}
\newcommand{\Val}{\mathcal{A}}
\newcommand{\Var}{\mathcal{V}}
\newcommand{\nf}[1]{{#1}\!\!\downarrow}
\newcommand{\vsim}{\sim}
\newcommand{\aequiv}{\mathrel{=_\alpha}}
\newcommand{\vsimv}[1]{\vsim_{#1}}
\newcommand{\tpos}[2]{#1\!\mid_{#2}}
\newcommand{\muvar}{\mathcal{Var}}

@@

The mapping $\Val$ is computed as follows.
%
For each equivalence class $\ec{x}$, we associate a fresh variable $\Var{\ec{x}}$ of the same type as $x$
%not occurring in $\tEc$,
and let $\Val\ec{x} = \Var{\ec{x}}$.
Thus, there are initially no constraints on the values for any equivalence class $\ec{x}$.
The mapping $\Val$ is then modified by applying the following rule exhaustively:
\[
\hbox{\(
\inferrule{
  \Var{\ec{x}} \in \FV( \Val )
  \quad
  \const{C}( t_1, \ldots, t_n ) \in \ec{x}
  \quad
  \const{C} \in \Ctr
}{
  \Val := \Val \{ \Var{\ec{x}} \mapsto \mu \Var{\ec{x}}.\; \const{C}( \Var{\ec{t_1}}, \ldots, \Var{\ec{t_n}} ) \}
}
\)
\rn{Update}}
\]
$\FV( \Val )$ denotes the union of the free variables occurring in $\Val$'s range.
It is straightforward to see that the height of terms produced as a result of this update
is bounded by the number of equivalence classes of $\Ec$,
and thus the construction of $\Val$ will terminate.

\begin{example}
Suppose that $\Ec$ contains distinct equivalence classes $\ec{w}$, $\ec{x}$, $\ec{y}$, and $\ec{z}$
where $\const{C}(z,x) \in \ec{y}$ and $\const{C}(w,y) \in \ec{x}$ for some $\const{C} \in \Ctr$.
A possible sequence of \rn{Update} rule applications is presented below.
\[\begin{array}{l@{\hspace{1em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l}
%\hline
%\text{Step} & \multicolumn{3}{l}{\text{Mapping }\Val}
%\MIDRULE
\text{1. Initial:}
& \{ & \ec{x} \mapsto \Var{\ec{x}}, & \ec{z} \mapsto \Var{\ec{z}}, & \\
&    & \ec{y} \mapsto \Var{\ec{y}}, & \ec{w} \mapsto \Var{\ec{w}} & \} \\[\jot]
\text{2. Update w.r.t.\ }\ec{x}\text{:}
& \{ & \ec{x} \mapsto \mu \Var{\ec{x}}. \const{C}( \Var{\ec{w}}, \Var{\ec{y}} ), & \ec{z} \mapsto \Var{\ec{z}}, & \\
&    & \ec{y} \mapsto \Var{\ec{y}}, & \ec{w} \mapsto \Var{\ec{w}} & \} \\[\jot]
\text{3. Update w.r.t.\ }\ec{y}\text{:}
& \{ & \ec{x} \mapsto \mu \Var{\ec{x}}. \const{C}( \Var{\ec{w}}, \mu \Var{\ec{y}}. \const{C}( \Var{\ec{z}}, \Var{\ec{x}} ) ), & \ec{z} \mapsto \Var{\ec{z}}, & \\
&    & \ec{y} \mapsto \mu \Var{\ec{y}}. \const{C}( \Var{\ec{z}}, \Var{\ec{x}} ), & \ec{w} \mapsto \Var{\ec{w}} & \} \\[\jot]
\text{4. Update w.r.t.\ }\ec{x}\text{:}
& \{ & \ec{x} \mapsto \mu \Var{\ec{x}}. \const{C}( \Var{\ec{w}}, \mu \Var{\ec{y}}. \const{C}( \Var{\ec{z}}, \Var{\ec{x}} ) ), & \ec{z} \mapsto \Var{\ec{z}}, & \\
&    & \ec{y} \mapsto \mu \Var{\ec{y}}. \const{C}( \Var{\ec{z}}, \mu \Var{\ec{x}}. \const{C}( \Var{\ec{w}}, \Var{\ec{y}} ) ), & \ec{w} \mapsto \Var{\ec{w}} & \}
%\hline
\end{array}\]
%The first column shows the equivalence class that was considered;
%the second column shows the content of $\Val$.
This indicates that the values for $x$ (resp. $y$) in models of $\Ec$
are of the form $\const{C}( \Var{\ec{w}}, \const{C}( \Var{\ec{z}}, \ldots ))$ (resp. $\const{C}( \Var{\ec{z}}, \const{C}( \Var{\ec{w}}, \ldots ))$) 
for some value of $\Var{\ec{w}}$ and $\Var{\ec{z}}$.
\qed
\end{example}

@@

\begin{example}
Let $\const{C}$ be a constructor of type $\tau_1 \times \tau_2 \rightarrow \tau_2$,
and let $\Ec$ be the set $\{ x \teq \const{C}(y,\const{C}(y,x)) \}$.
After applying the rules in Figure~\ref{fig:cc-rules} to completion, the equivalence classes of $\Ec$ are
$\{ x, \const{C}(y,\const{C}(y,x)) \}$, $\{ \const{C}(y,x) \}$, and $\{ y \}$.
We then construct $\Val$ using the aforementioned phases, after which we obtain:
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}r@{\hspace{.5em}}l@{\hspace{.5em}}r@{\hspace{.5em}}l}
\Val \ec{x} & = & 
\mu \Var{\ec{x}}. & \const{C}( \Var{\ec{y}}, & \mu \Var{\ec{\const{C}(y,x)}}. & \const{C}( \Var{\ec{y}}, \Var{\ec{x}} ) ) \\
\Val \ec{\const{C}(y,x)} & = & 
\mu \Var{\ec{\const{C}(y,x)}}. & \const{C}( \Var{\ec{y}}, & \mu \Var{\ec{x}}. & \const{C}( \Var{\ec{y}}, \Var{\ec{\const{C}(y,x)}} ) )
\end{array}\]
Notice that $\Val \ec{x}$ and $\Val \ec{\const{C}(y,x)}$ are syntactically equivalent for the renaming
$\{ \Var{\ec{x}} \mapsto \Var{\ec{\const{C}(y,x)}}, \Var{\ec{\const{C}(y,x)}} \mapsto \Var{\ec{x}} \}$,
and thus our calculus may infer $x \teq \const{C}(y,x)$.
Intuitively, this equality holds, since the values of $x$ and $\const{C}( y, x )$ 
are equivalent for any value of $\Var{\ec{y}}$.
\qed
\end{example}

\begin{figure}[t]
\normalsize
\centering
\begin{tabular}{c}
\(
\inferrule{
  t, u : \tau \in \tEc
  \quad
  \tau \in \Codata,
  \mid \tau \mid = 1
}{
  \Ec := \Ec, t \teq u
}
\)
\rn{Singleton}
\\[5\jot]
\(
\inferrule{
  t : \tau \in \tEc 
  \quad 
  ( s( t ) \in \tEc, s \in \Sel^\tau ) 
  \text{ or } 
  ( \tau \in \Data, \tau \text{ finite } )
}{
  \Ec := \Ec, t \teq \const{C}_1( s^1_1( t ), \ldots, s^{a_1}_1( t ) ) \ror \ldots \ror \Ec := \Ec, t \teq \const{C}_n( s^1_n( t ), \ldots, s^{a_n}_n( t ) ) 
}
\)
\rn{Split} 
\end{tabular}
\caption{\,Derivation rules for branching%.  
%All pairs of terms whose type has cardinality 1 are entailed to be equal (the degenerate case).
%Constructors must be assigned for all terms $t$ if has a selector is applied to it, or if $t$ has finite datatype type.
}
\label{fig:split-rule}
\end{figure}

\paragraph{Phase 3: Branching.}
If neither a rule from Figure \ref{fig:cc-rules}~or~\ref{fig:ab-rules} applies to $\Ec$, 
the rules from Figure~\ref{fig:split-rule} are applied.
The rule \rn{Singleton} handles the case when there are two terms from $\tEc$ having a co-recursive singleton type,
in which case they must be equivalent.
Otherwise, in the case there is a selector term applied to $t$, or if $t$ has a finite datatype type, 
then the equivalence class of $t$ must contain an application of one of the constructors of $\tau$ (where $\Ctr^\tau = \{ C_1, \ldots C_n \}$), 
which is enforced by the rule \rn{Split}.

@@

\begin{example}
Let $\tau$ be a type in $\Codata$ with three constructors $\const{A}$, $\const{B}$ and $\const{C}$ each having type $\tau \rightarrow \tau$.
Let $\Ec$ be the set consisting of equalities
$\{$
$y_1 \teq \const{A}( y_2 )$,
$y_2 \teq \const{B}( x )$,
$z_1 \teq \const{C}( z_2 )$,
$z_2 \teq \const{A}( z_1 )$,
$w_1 \teq \const{A}( x )$,
$w_2 \teq \const{C}( w_3 )$,
$x \not\teq y_2$,
$\}$.
After running the calculus to saturation on $\Ec$, the mapping $\Val$ contains:
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{2.0em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l}
\Val \ec{y_1} & = & \mu \Var{\ec{y_1}}. & \const{A}( & \mu \Var{\ec{y_2}}. & \const{B}( \Var{\ec{x}} ) ) &
\Val \ec{w_1} & = & \mu \Var{\ec{w_1}}. & \const{A}( \Var{\ec{x}} )      & \\ 
\Val \ec{y_2} & = & \mu \Var{\ec{y_2}}. & \multicolumn{2}{l}{\const{B}( \Var{\ec{x}} )}      & &
\Val \ec{w_2} & = & \mu \Var{\ec{w_2}}. & \const{C}( \Var{\ec{w_3}} ) \\
\Val \ec{z_1} & = & \mu \Var{\ec{z_1}}. & \const{C}( & \mu \Var{\ec{z_2}}. & \const{A}( \Var{\ec{z_1}} ) ) &
\Val \ec{w_3} & = & \Var{\ec{w_3}} \\
\Val \ec{z_2} & = & \mu \Var{\ec{z_2}}. & \const{A}( & \mu \Var{\ec{z_1}}. & \const{C}( \Var{\ec{z_2}} ) ) & 
\Val \ec{x} & = & \Var{\ec{x}} \\
\end{array}\]

To construct a completion of $\Val$, we must choose values for the free variables of $\Val$, $\Var{\ec{x}}$ and $\Var{\ec{w_3}}$.
The content of the set $\mathcal{T}^\M_{\Var{\ec{x}}}( \Val )$ is $\alpha$-equivalent to:
\[%\begin{array}{c}
\{
\mu x. \const{A}( \mu y. \const{B}( x )),
\mu x. \const{A}( \mu y. \const{B}( y )),
\mu x. \const{B}( x ),
\mu x. \const{C}( \mu y. \const{A}( x )),
\mu x. \const{A}( \mu y. \const{C}( x )),
\mu x. \const{A}( x )
\}
%\end{array}
\]
Now, consider a model $\M$ that interprets variables in $\Ec$ based on the mapping $\Val$,
for instance, $x$ is interpretted as $\Val \ec{x}$, $y_1$ as $\Val \ec{y_1}$ and so on.
Assigning a value for $\Var{\ec{x}}$ that is $\alpha$-equivalent to a value in $\mathcal{T}^\M_{\Var{\ec{x}}}( \Val )$
may cause values in the range of $\Val$ to become $\alpha$-equivalent, 
which in turn may cause $\M$ to be inconsistent.
For instance, say we assign $\mu \Var{\ec{x}}. \const{B}( \Var{\ec{x}} )$ for $\Var{\ec{x}}$.
After this substitution, $\Val \ec{y_2}$ is $\mu \Var{\ec{y_2}}. \const{B}( \mu \Var{\ec{x}}. \const{B}( \Var{\ec{x}} ) )$, 
which has normal form $\mu \Var{\ec{y_2}}. \const{B}( \Var{\ec{y_2}} )$,
which is $\alpha$-equivalent to $\mu \Var{\ec{x}}. \const{B}( \Var{\ec{x}} )$.
However, this contradicts the disequality $x \not\teq y_2$ in $\Ec$.
Inversely, if the value assigned to $\Var{\ec{x}}$ is not $\alpha$-equivalent to any term in $\mathcal{T}^\M_{\Var{\ec{x}}}( \Val )$,
then the values in the range of $\Val$ remain $\alpha$-disequivalent.
For instance, we may assign values such as 
$\mu \Var{\ec{x}}. \const{C}( \Var{\ec{x}} )$,
$\mu \Var{\ec{x}}. \const{C}( \mu x_1. \const{A}( x_1) )$, and
$\mu \Var{\ec{x}}. \const{B}( \mu x_1. \const{A}( \Var{\ec{x}} ) )$
to $\Var{\ec{x}}$.
Notice that legal substitutions for $\Var{\ec{x}}$ may cause the range of $\Val$ to contain non-normal terms,
as in after applying the third substitution mentioned above,
$\Val \ec{w_1}$ is $\mu \Var{\ec{w_1}}. \const{A}( \mu \Var{\ec{x}}. \const{B}( \mu x_1. \const{A}( \Var{\ec{x}} ) ) )$.
$\Box$
\end{example}

@@

\begin{lemma} 
\label{lem:model-completion}
If $\Val$ is constructed for a saturated $\Ec$,
and $\Val^\ast$ is a completion of $\Val$ for normal model $\M$, then:
\begin{enumerate}
\item $\Val^\ast \ec{x}$ is bisimilar with a value in $\M( \tau )$, for all $\ec{x}$ of type $\tau$ of $\Ec$,
\item
If $\const{C}( t_1, \ldots, t_n ) \in \ec{x}$,
then $\Val^\ast \ec{x} = \mu \Var{\ec{x}}. \const{C}( w_1, \ldots, w_n )$
and $\Val^\ast \ec{x} \interp{ w_i } = \Val^\ast \ec{t_i}$ for $i = 1, \ldots, n$.
%and $\muvar( w_i ) = \Var{\ec{t_i}}$ for $i = 1, \ldots, n$,
%\item
%$t_1 \interp{ t_2 } = \Val^\ast \ec{x}$
%for all $t_1$ with subterm $t_2$ in the range of $\Val^\ast$ where $\muvar(t_2)=\Var{\ec{x}}$,
\item 
$\Val^\ast \ec{x} \vsim \Val^\ast \ec{y}$ if and only if $\ec{x} = \ec{y}$.
\end{enumerate}
\end{lemma}
\begin{proof}
To show 1, we first show that $\Val^\ast$ contains no free variables.
Assume by contradiction that $\Val^\ast$ contained a free variable $\Var{\ec{y}}$ for some $\ec{y}$ of type $\tau$.
Then it must be the case that $\ec{y}$ does not contain a constructor term,
or else $\Var{\ec{y}}$ would not occur as a free variable in $\Val$.
Consider the case when $\tau$ is finite.
By assumption, $\tau \not\in \Nondata$.
Since \rn{Split} does not apply to $\Ec$, we have $\tau \not\in \Data$.
If $\tau \in \Codata$, then $\tau$ is corecursive by assumption, and by Lemma~\ref{lem:corecursive-singletons},
the cardinality of $\tau$ must be one.
Since \rn{Singleton} does not apply to $\Ec$,
there is only one equivalence class of type $\tau$ in $\Ec$,
and thus there are no terms in $\mathcal{T}^\M_x(\Val^\ast)$ of type $\tau$.
This is a contradiction, since our model completion may assign the value in the domain of $\tau$ to $\Var{\ec{x}}$.
Now, consider the case when $\tau$ is infinite.
This is also a contradiction, 
since there are only a finite number of closed terms in $\mathcal{T}^\M_x(\Val^\ast)$,
and thus our model completion can assign a value not occurring in $\mathcal{T}^\M_x(\Val^\ast)$ to $\Var{\ec{y}}$.
By construction, $\Val^\ast \ec{x}$ is normal.
Since \rn{Acyclic} does not apply, $\Val \ec{x}$ is acyclic when $\tau \in \Data$.
Moreover, our construction of $\Val^\ast$ applies substitutions of the form 
$\{ y \mapsto t \}$, where $t$ is acyclic when the type of $y$ is not a codatatype.
Thus, $\Val^\ast \ec{x}$ is acyclic when $\tau \not\in \Codata$.
Therefore, by definition, $\Val^\ast \ec{x}$ is bisimilar with a value in $\M( \tau )$.

>>>>

%\renewcommand{\vec}[1]{\mathbf{#1}}
\renewcommand{\vec}[1]{\bar #1}
\newcommand{\Ec}{\mathsf{E}}
\newcommand{\Fc}{\mathsf{F}}
\newcommand{\Gc}{\mathsf{G}}
\newcommand{\Ac}{\mathsf{A}}
\newcommand{\Dc}{\mathsf{D}}
\newcommand{\Rc}{\mathsf{R}}
\newcommand{\vrange}{\mathsf{range}}
\newcommand{\vdom}{\mathsf{dom}}
\newcommand{\tEc}{\mathcal{T}(\Ec)}
\newcommand{\tcEc}{\mathcal{T}^\ast(\Ec)}
\newcommand{\rn}[1]{\textsf{\small #1}}
\newcommand{\cvc}{\textsc{cvc}{\small 4}\xspace}
\newcommand{\teq}{\approx}
\newcommand{\tneq}{\not\teq}
\newcommand{\rem}[1]{\textcolor{red}{[#1]}}
%\newcommand{\is}[1]{is\text{-} #1}
\newcommand{\is}[1]{\const{is#1}}
\newcommand{\ror}{\quad \parallel \quad}
\newcommand{\tpath}[2]{\mathcal{P}_{ #2 \rightarrow \_ }( #1 )}
\newcommand{\ttpath}[3]{\mathcal{P}_{ #2 \rightarrow #3 }( #1 )}
\newcommand{\interp}[1]{[\![ #1 ]\!]}
\newcommand{\dpath}[3]{\delta^{#1}_{#2}( #3 )}
\newcommand{\ec}[1]{[#1]}
\newcommand{\M}{\mathcal{M}}
\newcommand{\Val}{\mathcal{A}}
\newcommand{\Var}{\mathcal{V}}
%\newcommand{\Varec}[1]{\Var{\ec{#1}}}
\newcommand{\Varec}[1]{\widetilde{#1}}
\newcommand{\nf}[1]{{#1}\!\!\downarrow}
\newcommand{\vsim}{\sim}
\newcommand{\aequiv}{\mathrel{=_\alpha}}
\newcommand{\vsimv}[1]{\vsim_{#1}}
\newcommand{\tpos}[2]{#1\!\mid_{#2}}
\newcommand{\muvar}{\mathcal{Var}}

@@

The mapping $\Val$ is computed as follows.
%
For each equivalence class $\ec{x}$, we associate a fresh variable ${\Varec{x}}$ of the same type as $x$
%not occurring in $\tEc$,
and let $\Val\ec{x} = {\Varec{x}}$.
Thus, there are initially no constraints on the values for any equivalence class $\ec{x}$.
The mapping $\Val$ is then modified by applying the following rule exhaustively:
\[
\hbox{\(
\inferrule{
  {\Varec{x}} \in \FV( \Val )
  \quad
  \const{C}( t_1, \ldots, t_n ) \in \ec{x}
  \quad
  \const{C} \in \Ctr
}{
  \Val := \Val \{ {\Varec{x}} \mapsto \mu {\Varec{x}}.\; \const{C}( {\Varec{t_1}}, \ldots, {\Varec{t_n}} ) \}
}
\)
\rn{Update}}
\]
$\FV( \Val )$ denotes the union of the free variables occurring in $\Val$'s range.
It is straightforward to see that the height of terms produced as a result of this update
is bounded by the number of equivalence classes of $\Ec$,
and thus the construction of $\Val$ will terminate.

\begin{example}
Suppose that $\Ec$ contains distinct equivalence classes $\ec{w}$, $\ec{x}$, $\ec{y}$, and $\ec{z}$
where $\const{C}(z,x) \in \ec{y}$ and $\const{C}(w,y) \in \ec{x}$ for some $\const{C} \in \Ctr$.
A possible sequence of \rn{Update} rule applications is presented below, omitting
trivial entries such as $\ec{w} \mapsto {\Varec{w}}$.
%
\begin{enumerate}
\item \noindent\rlap{Initial state:}\phantom{Update w.r.t.\ $\ec{x}$:\enskip}$\{  \}$
\item \noindent\rlap{Update w.r.t.\ $\ec{x}$:}\phantom{Update w.r.t.\ $\ec{x}$:\enskip}$\{\vthinspace  \ec{x} \mapsto \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\, {\Varec{y}} ) \}$
\item \noindent\rlap{Update w.r.t.\ $\ec{y}$:}\phantom{Update w.r.t.\ $\ec{x}$:\enskip}$\{\vthinspace  \ec{x} \mapsto \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\; \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) ),\;
  \ec{y} \mapsto \mu {\Varec{y}}. \const{C}( {\Varec{z}},\, {\Varec{x}} ) \vthinspace\}$
\item \noindent\rlap{Update w.r.t.\ $\ec{x}$:}\phantom{Update w.r.t.\ $\ec{x}$:\enskip}$\{\vthinspace  \ec{x} \mapsto \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\; \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) ),\;
  \ec{y} \mapsto \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\; \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\, {\Varec{y}} ) ) \vthinspace\}$
\end{enumerate}
%
This indicates that the values for $x$ and $y$ in models of $\Ec$
are of the form $\const{C}( {\Varec{w}}, \const{C}( {\Varec{z}}, \ldots ))$
and $\const{C}( {\Varec{z}}, \const{C}( {\Varec{w}}, \ldots ))$,
respectively, for some value of ${\Varec{w}}$ and ${\Varec{z}}$.
\end{example}

@@

\begin{example}
Let $\const{C}$ be a constructor of type $\tau_1 \times \tau_2 \rightarrow \tau_2$,
and let $\Ec$ be the set $\{ x \teq \const{C}(y,\const{C}(y,x)) \}$.
After applying the rules in Figure~\ref{fig:cc-rules} to completion, the equivalence classes of $\Ec$ are
$\{ x, \const{C}(y,\const{C}(y,x)) \}$, $\{ \const{C}(y,x) \}$, and $\{ y \}$.
We then construct $\Val$ using the aforementioned phases, after which we obtain:
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}r@{\hspace{.5em}}l@{\hspace{.5em}}r@{\hspace{.5em}}l}
\Val \ec{x} & = & 
\mu {\Varec{x}}. & \const{C}( {\Varec{y}}, & \mu {\Varec{\const{C}(y,x)}}. & \const{C}( {\Varec{y}}, {\Varec{x}} ) ) \\
\Val \ec{\const{C}(y,x)} & = & 
\mu {\Varec{\const{C}(y,x)}}. & \const{C}( {\Varec{y}}, & \mu {\Varec{x}}. & \const{C}( {\Varec{y}}, {\Varec{\const{C}(y,x)}} ) )
\end{array}\]
Notice that $\Val \ec{x}$ and $\Val \ec{\const{C}(y,x)}$ are syntactically equivalent for the renaming
$\{ {\Varec{x}} \mapsto {\Varec{\const{C}(y,x)}}, {\Varec{\const{C}(y,x)}} \mapsto {\Varec{x}} \}$,
and thus our calculus may infer $x \teq \const{C}(y,x)$.
Intuitively, this equality holds, since the values of $x$ and $\const{C}( y, x )$ 
are equivalent for any value of ${\Varec{y}}$.
\qed
\end{example}

\begin{figure}[t]
\normalsize
\centering
\begin{tabular}{c}
\(
\inferrule{
  \tau \in \Codata
  \quad
  \left|\tau\right| = 1
  \quad
  t, u : \tau \in \tEc
}{
  \Ec := \Ec,\: t \teq u
}
\)
\rn{Singleton}
\\[5\jot]
\(
\inferrule{
  t : \tau \in \tEc 
  \quad 
  ( s( t ) \in \tEc, s \in \Sel^\tau ) 
  \text{ or } 
  ( \tau \in \Data, \tau \text{ finite } )
}{
  \Ec := \Ec,\: t \teq \const{C}_1( s^1_1( t ), \ldots, s^{a_1}_1( t ) ) \ror \cdots \ror \Ec := \Ec, t \teq \const{C}_n( s^1_n( t ), \ldots, s^{a_n}_n( t ) ) 
}
\)
\rn{Split} 
\end{tabular}
\caption{\,Derivation rules for branching%.  
%All pairs of terms whose type has cardinality 1 are entailed to be equal (the degenerate case).
%Constructors must be assigned for all terms $t$ if has a selector is applied to it, or if $t$ has finite datatype type.
}
\label{fig:split-rule}
\end{figure}

\paragraph{Phase 3: Branching.}
If neither a rule from Figure \ref{fig:cc-rules}~or~\ref{fig:ab-rules} applies to $\Ec$, 
the rules from Figure~\ref{fig:split-rule} are applied.
The \rn{Singleton} rule handles the case when there are two terms from $\tEc$ having a co-recursive singleton type,
in which case they must be equivalent.
Otherwise, in the case there is a selector term applied to $t$, or if $t$ has a finite datatype type, 
then the equivalence class of $t$ must contain an application of one of the constructors of $\tau$ (where $\Ctr^\tau = \{ C_1, \ldots C_n \}$), 
which is enforced by the rule \rn{Split}.

@@

\begin{example}
Let $\tau$ be a type in $\Codata$ with three constructors $\const{A}$, $\const{B}$ and $\const{C}$ each having type $\tau \rightarrow \tau$.
Let $\Ec$ be the set consisting of equalities
$\{$
$y_1 \teq \const{A}( y_2 )$,
$y_2 \teq \const{B}( x )$,
$z_1 \teq \const{C}( z_2 )$,
$z_2 \teq \const{A}( z_1 )$,
$w_1 \teq \const{A}( x )$,
$w_2 \teq \const{C}( w_3 )$,
$x \not\teq y_2$,
$\}$.
After running the calculus to saturation on $\Ec$, the mapping $\Val$ contains:
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{2.0em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l}
\Val \ec{y_1} & = & \mu {\Varec{y_1}}. & \const{A}( & \mu {\Varec{y_2}}. & \const{B}( {\Varec{x}} ) ) &
\Val \ec{w_1} & = & \mu {\Varec{w_1}}. & \const{A}( {\Varec{x}} )      & \\ 
\Val \ec{y_2} & = & \mu {\Varec{y_2}}. & \multicolumn{2}{l}{\const{B}( {\Varec{x}} )}      & &
\Val \ec{w_2} & = & \mu {\Varec{w_2}}. & \const{C}( {\Varec{w_3}} ) \\
\Val \ec{z_1} & = & \mu {\Varec{z_1}}. & \const{C}( & \mu {\Varec{z_2}}. & \const{A}( {\Varec{z_1}} ) ) &
\Val \ec{w_3} & = & {\Varec{w_3}} \\
\Val \ec{z_2} & = & \mu {\Varec{z_2}}. & \const{A}( & \mu {\Varec{z_1}}. & \const{C}( {\Varec{z_2}} ) ) & 
\Val \ec{x} & = & {\Varec{x}} \\
\end{array}\]

To construct a completion of $\Val$, we must choose values for the free variables of $\Val$, ${\Varec{x}}$ and ${\Varec{w_3}}$.
The content of the set $\mathcal{T}^\M_{{\Varec{x}}}( \Val )$ is $\alpha$-equivalent to:
\[%\begin{array}{c}
\{
\mu x. \const{A}( \mu y. \const{B}( x )),
\mu x. \const{A}( \mu y. \const{B}( y )),
\mu x. \const{B}( x ),
\mu x. \const{C}( \mu y. \const{A}( x )),
\mu x. \const{A}( \mu y. \const{C}( x )),
\mu x. \const{A}( x )
\}
%\end{array}
\]
Now, consider a model $\M$ that interprets variables in $\Ec$ based on the mapping $\Val$,
for instance, $x$ is interpretted as $\Val \ec{x}$, $y_1$ as $\Val \ec{y_1}$ and so on.
Assigning a value for ${\Varec{x}}$ that is $\alpha$-equivalent to a value in $\mathcal{T}^\M_{{\Varec{x}}}( \Val )$
may cause values in the range of $\Val$ to become $\alpha$-equivalent, 
which in turn may cause $\M$ to be inconsistent.
For instance, say we assign $\mu {\Varec{x}}. \const{B}( {\Varec{x}} )$ for ${\Varec{x}}$.
After this substitution, $\Val \ec{y_2}$ is $\mu {\Varec{y_2}}. \const{B}( \mu {\Varec{x}}. \const{B}( {\Varec{x}} ) )$, 
which has normal form $\mu {\Varec{y_2}}. \const{B}( {\Varec{y_2}} )$,
which is $\alpha$-equivalent to $\mu {\Varec{x}}. \const{B}( {\Varec{x}} )$.
However, this contradicts the disequality $x \not\teq y_2$ in $\Ec$.
Inversely, if the value assigned to ${\Varec{x}}$ is not $\alpha$-equivalent to any term in $\mathcal{T}^\M_{{\Varec{x}}}( \Val )$,
then the values in the range of $\Val$ remain $\alpha$-disequivalent.
For instance, we may assign values such as 
$\mu {\Varec{x}}. \const{C}( {\Varec{x}} )$,
$\mu {\Varec{x}}. \const{C}( \mu x_1. \const{A}( x_1) )$, and
$\mu {\Varec{x}}. \const{B}( \mu x_1. \const{A}( {\Varec{x}} ) )$
to ${\Varec{x}}$.
Notice that legal substitutions for ${\Varec{x}}$ may cause the range of $\Val$ to contain non-normal terms,
as in after applying the third substitution mentioned above,
$\Val \ec{w_1}$ is $\mu {\Varec{w_1}}. \const{A}( \mu {\Varec{x}}. \const{B}( \mu x_1. \const{A}( {\Varec{x}} ) ) )$.
$\Box$
\end{example}

@@

\begin{lemma} 
\label{lem:model-completion}
If $\Val$ is constructed for a saturated $\Ec$,
and $\Val^\ast$ is a completion of $\Val$ for normal model $\M$, then:
\begin{enumerate}
\item $\Val^\ast \ec{x}$ is bisimilar with a value in $\M( \tau )$, for all $\ec{x}$ of type $\tau$ of $\Ec$,
\item
If $\const{C}( t_1, \ldots, t_n ) \in \ec{x}$,
then $\Val^\ast \ec{x} = \mu {\Varec{x}}. \const{C}( w_1, \ldots, w_n )$
and $\Val^\ast \ec{x} \interp{ w_i } = \Val^\ast \ec{t_i}$ for $i = 1, \ldots, n$.
%and $\muvar( w_i ) = {\Varec{t_i}}$ for $i = 1, \ldots, n$,
%\item
%$t_1 \interp{ t_2 } = \Val^\ast \ec{x}$
%for all $t_1$ with subterm $t_2$ in the range of $\Val^\ast$ where $\muvar(t_2)={\Varec{x}}$,
\item 
$\Val^\ast \ec{x} \vsim \Val^\ast \ec{y}$ if and only if $\ec{x} = \ec{y}$.
\end{enumerate}
\end{lemma}
\begin{proof}
To show 1, we first show that $\Val^\ast$ contains no free variables.
Assume by contradiction that $\Val^\ast$ contained a free variable ${\Varec{y}}$ for some $\ec{y}$ of type $\tau$.
Then it must be the case that $\ec{y}$ does not contain a constructor term,
or else ${\Varec{y}}$ would not occur as a free variable in $\Val$.
Consider the case when $\tau$ is finite.
By assumption, $\tau \not\in \Nondata$.
Since \rn{Split} does not apply to $\Ec$, we have $\tau \not\in \Data$.
If $\tau \in \Codata$, then $\tau$ is corecursive by assumption, and by Lemma~\ref{lem:corecursive-singletons},
the cardinality of $\tau$ must be one.
Since \rn{Singleton} does not apply to $\Ec$,
there is only one equivalence class of type $\tau$ in $\Ec$,
and thus there are no terms in $\mathcal{T}^\M_x(\Val^\ast)$ of type $\tau$.
This is a contradiction, since our model completion may assign the value in the domain of $\tau$ to ${\Varec{x}}$.
Now, consider the case when $\tau$ is infinite.
This is also a contradiction, 
since there are only a finite number of closed terms in $\mathcal{T}^\M_x(\Val^\ast)$,
and thus our model completion can assign a value not occurring in $\mathcal{T}^\M_x(\Val^\ast)$ to ${\Varec{y}}$.
By construction, $\Val^\ast \ec{x}$ is normal.
Since \rn{Acyclic} does not apply, $\Val \ec{x}$ is acyclic when $\tau \in \Data$.
Moreover, our construction of $\Val^\ast$ applies substitutions of the form 
$\{ y \mapsto t \}$, where $t$ is acyclic when the type of $y$ is not a codatatype.
Thus, $\Val^\ast \ec{x}$ is acyclic when $\tau \not\in \Codata$.
Therefore, by definition, $\Val^\ast \ec{x}$ is bisimilar with a value in $\M( \tau )$.
